```{r}
library(tidyverse)
library(lubridate)
df <- read.csv("Sample - Superstore.csv")
colnames(df)
head(df)
```

# Question 1: How does sales performance vary over time?

### Identify which months consistently show sales peaks and find any unusual dips in the trend line

```{r}
df <- df %>%
  mutate(
    Order.Date = case_when(
      str_detect(Order.Date, "/") ~ parse_date_time(Order.Date, orders = c("mdy", "dmy")),  # Handles both formats
      TRUE ~ as.Date(Order.Date)  # Keep it unchanged if already in Date format
    )
  )
```

```{r}
df <- df %>%
  mutate(Month = format(Order.Date, "%m"))  # Ensures "01"-"12"

monthly_sales <- df %>%
  group_by(Month) %>%
  summarise(Total_Sales = sum(Sales, na.rm = TRUE)) %>%
  arrange(Month)

# Ensure all 12 months appear
all_months <- tibble(Month = sprintf("%02d", 1:12))

monthly_sales <- full_join(all_months, monthly_sales, by = "Month") %>%
  mutate(Total_Sales = replace_na(Total_Sales, 0))

# Plot it
ggplot(monthly_sales, aes(x = as.numeric(Month), y = Total_Sales)) +
  geom_line(color = "cyan", linewidth = 1) +      
  geom_point(color = "red", size = 2) +           
  geom_smooth(method = "loess", formula = y ~ x, color = "orange", se = FALSE) +  
  scale_x_continuous(breaks = 1:12, labels = month.abb) +  
  labs(title = "Total Sales by Month",
       x = "Month",
       y = "Total Sales") +
  theme_minimal()
```

```{r}
print(monthly_sales)
```

### Calculate YOY growth rate for the most recent complete year

```{r}
df <- df %>%
  mutate(Year = format(Order.Date, "%Y"))  # Extract year

yearly_sales <- df %>%
  group_by(Year) %>%
  summarise(Total_Sales = sum(Sales, na.rm = TRUE)) %>%
  arrange(Year) %>%
  mutate(YOY_Growth = (Total_Sales / lag(Total_Sales) - 1) * 100)  # YOY formula

print(yearly_sales)
```

### Suggest potential reasons for seasonal patterns

From what we see in the graph, possible reasons for sales patterns:

📌 High Sales Months (Sep, Nov, Dec, maybe Mar/Oct):

November-December → Black Friday + Christmas shopping (retail boom)\
September-October → Back-to-school / Q4 business spending\
March → New product releases? Tax refunds? Q1 spending?

📌 Low Sales Months (Jan-Feb, Apr-Aug):

January-February → Post-holiday dip (people spent too much in Nov/Dec)\
April-August → Summer slowdown? Fewer business purchases?

------------------------------------------------------------------------

# Question 2: Which product categories have the best/worst profit margins?

### Identify which category has the thinnest margins

```{r}
category_margins <- df %>%
  group_by(Category) %>%
  summarise(
    Total_Sales = sum(Sales, na.rm = TRUE),
    Total_Profit = sum(Profit, na.rm = TRUE),
    Profit_Margin = (Total_Profit / Total_Sales) * 100
  ) %>%
  arrange(Profit_Margin)  # Sort from lowest to highest

print(category_margins)
```

### Calculate the profit margin difference between top and bottom categories

```{r}
max_margin <- max(category_margins$Profit_Margin, na.rm = TRUE)
min_margin <- min(category_margins$Profit_Margin, na.rm = TRUE)
margin_difference <- max_margin - min_margin

print(paste("Profit margin difference: ", round(margin_difference, 2), "%"))
```

### Suggest strategies to improve low-performing categories

📌 If margins are low due to discounts:

Reduce unnecessary discounts or promotions.\
Focus on premium product lines with higher markups.

📌 If margins are low due to high costs:

Negotiate better supplier deals.\
Optimize inventory management to reduce holding costs.

📌 If sales are high but profit is low:

Increase prices slightly if customers are price-insensitive.\
Cross-sell high-margin accessories or services with these products.

------------------------------------------------------------------------

# Question 3: How do regional performances compare?

### Identify which region has both high sales and high profitability and find any regions with negative profits

```{r}
region_performance <- df %>%
  group_by(Region) %>%
  summarise(
    Total_Sales = sum(Sales, na.rm = TRUE),
    Total_Profit = sum(Profit, na.rm = TRUE),
    Profit_Margin = (Total_Profit / Total_Sales) * 100
  ) %>%
  arrange(desc(Total_Sales))  # Sort by highest sales first

print(region_performance)
```

### Analyze if high sales always correlate with high profits

```{r}
ggplot(region_performance, aes(x = Total_Sales, y = Total_Profit, label = Region)) +
  geom_point(color = "blue", size = 4) +  
  geom_text(vjust = -0.5) +  
  labs(title = "Sales vs. Profit by Region", x = "Total Sales", y = "Total Profit") +
  theme_minimal()
```

### Propose regional-specific strategies based on findings

📌 If a region has high sales but low profit:

Reduce discounts or increase pricing in that region.\
Improve operational efficiency to cut costs.

📌 If a region has low sales but high profit margin:

Expand market reach through promotions or partnerships.\
Consider targeted marketing to increase volume.

📌 If a region is losing money (negative profit):

Investigate reasons (high discounts, expensive shipping, low demand).\
Adjust pricing, remove unprofitable products, or optimize logistics.

------------------------------------------------------------------------

# Question 4: What does customer segmentation reveal?

### Calculate percentage of customers in each segment

```{r}
# Find most recent order date
max_date <- max(df$Order.Date, na.rm = TRUE)

# Calculate RFM Metrics
rfm <- df %>%
  group_by(Customer.ID) %>%
  summarise(
    Recency = as.numeric(max_date - max(Order.Date)),  # Days since last purchase
    Frequency = n_distinct(Order.ID),  # Total number of purchases
    Monetary = sum(Sales, na.rm = TRUE)  # Total spent
  )

# Assign RFM Scores (1 = lowest, 5 = highest)
rfm <- rfm %>%
  mutate(
    R_Score = ntile(-Recency, 5),  # Recent buyers get high scores
    F_Score = ntile(Frequency, 5),  # Frequent buyers get high scores
    M_Score = ntile(Monetary, 5)  # High spenders get high scores
  ) %>%
  mutate(RFM_Score = R_Score * 100 + F_Score * 10 + M_Score)  # Combine into a single score

print(head(rfm))
```

```{r}
rfm <- rfm %>%
  mutate(Segment = case_when(
    RFM_Score >= 555 ~ "Best Customers",
    RFM_Score >= 450 ~ "Loyal Customers",
    RFM_Score >= 350 ~ "Potential Loyalists",
    RFM_Score >= 250 ~ "At Risk",
    TRUE ~ "Lost Customers"
  ))

# Count % of customers in each segment
segment_counts <- rfm %>%
  group_by(Segment) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = (Count / sum(Count)) * 100)

print(segment_counts)
```

### Identify which segment generates the most revenue

```{r}
df <- df %>%
  left_join(rfm, by = "Customer.ID")

segment_revenue <- df %>%
  inner_join(rfm, by = "Customer.ID") %>%
  group_by(Segment) %>%
  summarise(Total_Revenue = sum(Sales, na.rm = TRUE)) %>%
  arrange(desc(Total_Revenue))

print(segment_revenue)
```

### Strategies for Different Segments

📌 Retention Strategies for “At Risk” Customers:

Send personalized offers or discounts.\
Re-engagement emails (e.g., “We miss you! Here’s 10% off”).\
Check if they stopped buying due to pricing, product issues, or competition.

📌 Marketing Approaches for “High Spenders” (Best/Loyal Customers):

Offer exclusive deals, early access to new products.\
Loyalty programs to reward them.\
Encourage referrals (VIP referral bonuses).

------------------------------------------------------------------------

# Question 5: How does shipping mode affect profitability?

### Compare profit margins across shipping modes

```{r}
shipping_performance <- df %>%
  group_by(Ship.Mode) %>%
  summarise(
    Total_Sales = sum(Sales, na.rm = TRUE),
    Total_Profit = sum(Profit, na.rm = TRUE),
    Profit_Margin = (Total_Profit / Total_Sales) * 100
  ) %>%
  arrange(desc(Profit_Margin))  # Sort by highest profit margin

print(shipping_performance)
```

### Calculate profit per order for each shipping mode

```{r}
profit_per_order <- df %>%
  group_by(Ship.Mode) %>%
  summarise(
    Total_Orders = n_distinct(Order.ID),
    Total_Profit = sum(Profit, na.rm = TRUE),
    Profit_Per_Order = Total_Profit / Total_Orders
  ) %>%
  arrange(desc(Profit_Per_Order))

print(profit_per_order)
```

### Suggest optimal shipping strategy

📌 If a shipping mode has high sales but low profit margins:

Consider increasing shipping fees or adjusting pricing.\
Reduce unnecessary discounts on fast shipping.

📌 If a shipping mode is unprofitable:

Look at delivery costs—maybe it's too expensive for low-margin products.\
Set a minimum order value for free/fast shipping.

📌 If a shipping mode is highly profitable:

Encourage customers to use this mode (e.g., make it the default).\
Offer discounts for bulk orders to optimize shipping costs.
